---
title: "Optimizing EV Charging Infrastructure at California Cinemas"
author: "Conor McCaffery"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

**Live Dashboard:** [Click here to view the interactive EV Charging Infrastructure Mapping](https://conormccaffery.github.io/ev-infrastructure-analysis/)

The objective of this analysis is to identify highly profitable, localized monopolies for Level 2 Electric Vehicle (EV) charging investments in the California market. While EV adoption continues to scale rapidly—making up over 20% of new global car sales—public infrastructure has struggled to keep pace in dense areas.

Commercial movie theaters present a uniquely lucrative opportunity to bridge this gap. Because theaters guarantee a captive audience with a 2-to-3 hour dwell time, they perfectly align with the charging duration required for Level 2 EV stations. This project utilizes an ETL (Extract, Transform, Load) pipeline to aggregate three primary data sources by zip code: California EV registrations, existing alternative fueling stations, and commercial cinema locations.


## Methodology & Data Validation
Information for the California cinema dataset was sourced from Overpass Turbo. Because crowd-sourced spatial data often contains anomalies, manual data validation was a critical step in this pipeline. Initial data exploration revealed missing geographic tags for major flagship locations, such as a major local cinema in Santa Clara. Local market knowledge was applied to impute the missing zip code prior to SQL aggregation, ensuring high-value, high-traffic targets were accurately captured in the final model. CSV data containing EV registrations, existing charging stations, and theater coordinates were ingested into a local **SQLite data warehouse**



## Executive Summary
The spatial aggregation of DMV vehicle registrations and alternative fueling data reveals a stark infrastructure gap in California. While EV adoption is highly concentrated in specific zip codes, commercial entertainment venues—specifically cinemas, which offer the ideal 2-to-3 hour customer dwell time for Level 2 charging—are vastly underserved. This analysis identifies localized monopolies where early infrastructure investors can capture high-value, captive audiences.




```{r data-pipeline, results='hide'}
library(dplyr)

#Dataset 1: Alternative Fueling Stations Data 
ca_ev_data <- read.csv("alt_fuel_stations_ca.csv")
head(ca_ev_data)
colnames(ca_ev_data)


#cols_to_keep <- c("City","State","EV.Network","ZIP","EV.Pricing","Funding.Sources")
ca_df <- select(ca_ev_data,City,State,EV.Network,ZIP,EV.Pricing,Funding.Sources)
ca_df <- rename(ca_df,zip_code = ZIP)
```
```{r}
#Dataset 2: 2025 EV Registrations by Zip Code
ev_zip_data <- read.csv("vehicle-fuel-type-counts-2024.csv")
ev_zip_data <- select(ev_zip_data,-Date)
clean_ev_data <- subset(ev_zip_data, Fuel == 'Battery Electric' | Fuel == 'Plug-in Hybrid')
clean_ev_data <- rename(clean_ev_data,zip_code=ZIP.Code)
```
```{r}
# Dataset 3: Points of Interest
# Preparing to do a join on zip code
ca_cinemas <- read.delim("california_cinemas.csv",sep = "\t")
ca_cinemas <- ca_cinemas %>% rename(lat=1,long=2,zip_code=3)
ca_cinemas <- ca_cinemas %>% select(lat,long,zip_code,name) #dropping the amenity column

ca_cinemas <- ca_cinemas %>% mutate(zip_code = ifelse(grepl("Mercado",name, ignore.case = TRUE), "95054", zip_code ))
```

```{r}
# Database Architecture and SQL Execution
library(DBI)
library(RSQLite)
# Creating a temporary SQLite database 
con <- dbConnect(RSQLite::SQLite(), "ev_market_database.sqlite")

# Putting the three clean datasets into SQL database as tables
dbWriteTable(con, "chargers", ca_df,overwrite = TRUE)
dbWriteTable(con, "EVs", clean_ev_data, overwrite = TRUE)
dbWriteTable(con, "cinemas", ca_cinemas, overwrite = TRUE)

# Reading external SQL file in my PC
sql_query <- readLines("ev_market_aggregation.sql")
sql_query <- paste(sql_query, collapse = "\n")

ranked_opportunities <- dbGetQuery(con, sql_query)
dbDisconnect(con)
```

```{r}
head(ranked_opportunities)
```


The opportunity Score metric was calculated as 'Total EVs / (Existing Chargers + 1)'. The top three highest-value targets for immediate infrastructure expansion are:
1. **Orinda, East Bay (94563):** 3,130 registered EVs sharing zero public charging stations. A pure local monopoly,
2. **San Francisco - Castro (94114):** 2,662 registered EVs with zero existing public stations in a dense, parking-constrained neighborhood
3. **San Francisco - West Portal (94127):** 1,981 EVs with zero chargers, representing a massive suburban-urban foothold.
```{r}
head(ranked_opportunities,15)
```
## Analyze
The Opportunity Score metric highlights a clear mismatch between EV adoption and charging infrastructure across California. Some zip codes, such as 92677 with 14 existing public chargers, appear to be well-served relative to local EV demand, while others, particularly in the Bay Area, show severe shortfalls.
Notably, despite Los Angeles County having one of the highest total EV registration counts in the state, no LA zip codes appear among the top opportunity scores. This suggests charger deployment in LA has scaled roughly in line with demand, making the Bay Area's zero-competition zip codes a comparatively higher-priority target for new infrastructure investment.


## Strategic Mapping

```{r mapping, out.width='100%', echo=TRUE}
library(leaflet)
library(RColorBrewer)

cinema_map_data <- ca_cinemas %>% 
  filter(!is.na(lat) & !is.na(long) & zip_code != "") %>%
  left_join(ranked_opportunities, by = "zip_code")
  
cinema_map_data <- cinema_map_data %>% filter(lat >= 32.5 & lat <= 42.0, 
                                              long >= -124.5 & long <= -114.0)
# California Coordinates are positioned approximately between these values above.

# A source for the mapping error before cleaning comes from dataset 2 which tracks vehicle registrations, it might be common for people with California registrations to move to nearby states.
any(!startsWith(cinema_map_data$zip_code,"9"))
#California zip codes always start with 9
cinema_map_data <- cinema_map_data %>% filter(substr(zip_code,start = 1, stop = 1) == "9")

opportunity_score_palette <- colorNumeric(
  palette = "YlGn",
  domain = cinema_map_data$opportunity_score)

ev_strategy_map <- leaflet(data = cinema_map_data) %>% 
  # Map Layer 1
  addProviderTiles(providers$CartoDB.Positron) %>% 
  # Map Layer 2
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    fillColor = ~opportunity_score_palette(opportunity_score),
    color = "black",
    radius = 8,
    fillOpacity = 0.8,
    stroke = FALSE,
    
    # Map Layer 3
    popup = ~paste(
      "<b>Theater:</b>", name, "<br>",
      "<b>Zip Code:</b>", zip_code, "<br>",
      "<b>Total EVs:</b>", total_evs, "<br>",
      "<b>Existing Chargers:</b>", total_chargers, "<br>",
      "<b>Opportunity Score:</b>", round(opportunity_score, 0)
    )
  ) %>%
  # Map Layer 4
  addLegend(
    position = "bottomright",
    pal = opportunity_score_palette,
    values = ~opportunity_score,
    title = "EV Opportunity Score",
    opacity = 1
  )
```

```{r}
ev_strategy_map
```
## Key Findings
The Opportunity Score metric was engineered to measure market demand against available supply, calculated as Total EVs / (Existing Chargers + 1). The spatial mapping reveals that while EV adoption is heavily concentrated, infrastructure is severely lagging in specific pockets.

While the majority of California cinemas yield a baseline opportunity score of approximately 500, the algorithm identified extreme outliers concentrated heavily in the Bay Area (San Francisco and the East Bay). These locations represent pure, zero-competition monopolies for the theaters located within them.

## Strategic Recommendations
The top three highest-scoring zip codes should be prioritized for immediate site surveys and property management partnership outreach:

Orinda, East Bay (94563): The highest-value target in the state. There are 3,130 registered EVs sharing zero public charging stations. The local cinema holds a pure monopoly on EV-driving customers.

San Francisco - Castro (94114): 2,662 registered EVs with zero existing public stations in a dense, parking-constrained urban environment.

San Francisco - West Portal (94127): 1,981 EVs with zero chargers, representing a massive suburban-urban foothold.

Secondary strategic initiatives should include localized site assessments at these locations to evaluate physical parking lot capacity and electrical grid access prior to deployment.

## Limitations and Future Work
While the Opportunity Score effectively highlights zip-code-level disparities between EV adoption and charging infrastructure, several factors were outside the scope of this analysis. The score treats all chargers within a zip code as equally competitive regardless of their distance from a given cinema; a future iteration could use the existing latitude/longitude data to calculate a precise distance-to-nearest-charger for each theater. Additionally, charger counts do not distinguish between charging speeds or reflect real-time utilization, and cinema-level attendance data was unavailable, meaning all locations are currently weighted equally regardless of foot traffic.